Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Work in progress on (mainly) section 4.3.

More than a century of exploration and geological investigations of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the south-western and north-eastern margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.inpress},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

Independently of the choice of palaeo-temperature record\renote{
    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
    could possibly apply a much wilder choice of atmospheric forcing
    techniques, leading just as wild differences in model output
    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.inpress}.
There is patchy stratigraphical evidence for glaciations older than the Fraser
Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
timing are therefore still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
    \citet{Kleman.etal.2010, but I don't think that I have the time for this
    right now.}}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range (Fig. X). We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. In contrast, a well-developed network of glacial valleys west of
the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
that either our model does not perform reliably in this part of the domain or
the observed landscape pattern originates from older glacial cycles and
indicates an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Fig.~\ref{fig:timeseries}). The
6\,km-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,kyr, followed by a continuous deglaciation between
about 17 and 11\,kyr (Fig.~\ref{fig:deglacseries}, red dotted curve).
That driven by the GRIP record reaches maximal
volume at 19.1\,kyr, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,kyr, and a late
glacial re-advance between 12.6 and 11.6\,kyr (Fig.~\ref{fig:deglacseries},
blue dotted curve).

The possibly of a widespread late-glacial in
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
re-advance include alpine glacier moraines in the eastern Coast Mountains,
Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
further work is needed to constrain the timing of the late-glacial re-advance,
to seize its extent and to identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel)%
\footnote{Should we mention our work in progress with Brian and/or the Tuya
    Lake moraine?}.

Although the predicted timing of glacier retreat differs between the two
simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. After 12\,kyr, deglaciation proceeds
through an opening of the Liard lowland with radial margin retreat towards the
surrounding mountain ranges and inward flow towards the depression, in
compatibility with the regional melt water record \citep{Margold.etal.2013}.
Remaining ice continues its decay by retreating towards the Selwyn Mountains
and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
versus 9.4\,kyr in the GRIP simulation.

Once more, snapshots of modelled deglacial ice sheet configurations
(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}%
\footnote{I produced one figure for each run, but as you could expect from
    Fig.~\ref{fig:deglac}, Figs.~\ref{fig:deglacshots-grip}
    and~\ref{fig:deglacshots-epica} look very similar. One of them would
    probably be enough, or?}) provide a
poor medium for direct comparison with geomorphological evidence of the ice
retreat. Here, we compute a map of deglacial, basal flow directions immediately
preceding exposure of ice-free land or cessation of sliding upon the approach
of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
southern front of the ice sheet and in the Liard Lowland, the resulting map
show similarities with patterns of glacial lineations
(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~2]{Margold.etal.2013}).
However, such is not the case on the Interior Plateau of British Columbia,
where both simulations predicts negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), whereas glacial lineation maps indicate substantial
eastward flow.

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2014}.
On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the last glacial maximum, and may have remained intact throughout the
deglaciation.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
corresponding to documented extensive glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate past
climate evolution, and thus
can be seen as a first-order agreement between the model and geological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle, confirming previous inferences from
geological evidence preceding the last glacial maximum. However, our
simulations produce persistent ice cover on the Skeena mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the last glacial maximum re-advance, and
may have contributed to the spectacular glacial imprint of the Skeena Mountains
erosional landscape.

During deglaciation, none of the palaeo-temperature records used produces an
ideal agreement between model results and geological evidence. Although
the EPICA record yields to more realistic timing of the last glacial maximum
and early deglaciation, only the GRIP record produces a late-glacial re-advance
at places where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA record, and show a rapid deglaciation of the southern
half of the ice including rapid northwards retreat through the Interior
Plateau of central British Comumbia, followed by an opening of the Liard
Lowland, and final ice cap retreat of the ice margin towards the Selwyn and,
then, the Skeena Mountains, which host the last remnant of the ice sheet
during the early Holocene (10.9--9.5\,kyr).

These results strongly depend on our choices of ice-sheet model (PISM), surface
mass balance model (PDD) and climate forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, potentially leading to overestimated glacial
extent and volume in continental regions. Nevertheless, our results identify
the glacially-scrubbed Skeena Mountains as a central, key area to understanding
glacial dynamics of the Cordilleran ice sheet, highlighting the need for
further geological investigation of this region.

@@

% Acknowledgements
\begin{acknowledgements}
We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
providing constant help and development on PISM. This work was supported by the
Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot, and by NASA grant no.~NNX13AM16G
to E.~Bueler. Computer resources were provided by the Swedish National
Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
at the National Supercomputing Center (NSC).
\end{acknowledgements}

>>>>

More than a century of exploration and geological investigations of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the south-western and north-eastern margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

Independently of the choice of palaeo-temperature record\renote{
    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
    could possibly apply a much wilder choice of atmospheric forcing
    techniques, leading just as wild differences in model output
    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.2014}.
There is patchy stratigraphical evidence for glaciations older than the Fraser
Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
timing are therefore still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
    \citet{Kleman.etal.2010}, but I don't think that I have the time for this
    right now.}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range (Fig. X). We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. In contrast, a well-developed network of glacial valleys west of
the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
that either our model does not perform reliably in this part of the domain or
the observed landscape pattern originates from older glacial cycles and
indicates an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

\todo{add 1 sentence to connect to the next paragraph: that is easy:D}

The timing of peak ice volume during the last glacial maximum and the pacing of
deglaciation depend cirticially on the choice of atmospheric forcing
(Figs.~\ref{fig:timeseries} and~\ref{fig:deglacseries}). Adopting the EPICA
atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
curves). The simulation driven by the GRIP palaeo-temperature record yields
peak ice volume at 19.1\,ka and a deglaciation interupted by two standstills or
readvances until 9.4\,ka. The first interuption occurs between 16.6 and
14.5\,ka, and the second between 12.6 and 11.6\,ka
(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
runs, while similar in overall timing compared to runs with other climate
drivers, differ in detail in that the EPICA depicts peak glaciation almost
2\,ka later and shows a faster uninterrupted deglaciation which yields ice-free
conditions more than 1\,ka earlier.

\todo{From the timing of max glaciation and the duration of the deglaciation,
is there any evidence to prefer one model output above the other? If so, state
that, if not, state that too. The behavior of both models is perhaps reasonable
within the uncertainty envelope that the field evidence provides...?}

The possibility of late glacial readvances in the North American Cordillera has
been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
Squamish valleys off the southern tip of the Coast Mountains, indicates
consecutive glacier maximal (or stillstands while in overall retreat), one of
which one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area of northern Rocky
Mountains, sharp-crested moraines indicate a late-glacial readvance of alpine
glaciers and their interaction with the main body of the decaying Cordilleran
ice sheet \citep{Lakeman.etal.2008}. Additional evidence for late-glacial
alpine glacier readvances includes moraines in the eastern Coast Mountains,
Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
Menounos.etal.2008}. Although further work is needed to constrain the timing of
the late-glacial readvance, to assess its extents and geographical
distribution, and to identify the potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial readvance in the Coast
Mountains, Rocky Mountains and the Finlay River area, corresponding to where it
has been identified in the geological record (Fig.~\ref{fig:deglac}, left
panel). In contrast the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a highly restricted glacial readvance
(Fig.~\ref{fig:deglac}, right panel).

\todo{Note that most of the evidence for re-advance is for small glaciers,
      whereas the model output indicates that there was still a big body of ice
      at that time, as in \citet{Lakeman.etal.2008}.}

Predicted patterns of ice sheet retreat are relatively consistent between the
two simulations (Fig.~\ref{fig:deglac}). The southern sector of the modelling
domain, including the Puget Lowland, the Coast and the Rocky mountains, and the
Interior Plateau of British Columbia, becomes completely deglaciated by 12\,ka,
whereas a significant ice cover remains over the Skeena, Selwyn, MacKenzie,
Wrangell and Saint-Elias mountains in the northern sector of the modelling
domain. After 12\,ka, deglaciation continues to proceed across the Liard
Lowland with a radial ice margin retreat towards the surrounding mountain
ranges, and inward flow towards the depression, consistent with the regional
melt water record of the last deglaciation \citep{Margold.etal.2013}
\todo{rephrase with ``unzip''?}. Remaining ice continues to decay by retreating
towards the Selwyn and Skeena mountains. The last remnants of the Cordilleran
ice sheet finally disappears from the Skeena mountains around 10.6\,ka (EPICA)
and 9.4\,ka (GRIP).

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelop
\citep[terminology from]{Kleman.etal.2006}, that is that these were formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present a map of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside the cold-base retreating margin (Fig.~\ref{fig:lastflow})
\todo{description of the general trends that the map shows}. Patterns of
glacial lineation formed in the northern and southern sectors of the
Cordilleran ice sheet and in the Liard Lowland (\citealp{Prest.etal.1968};
\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010};
\citealp[Fig.~2]{Margold.etal.2013}), similarities with the patterns of
deglacial ice flow from modelling (Fig.~\ref{fig:lastflow}). However, this is
not the case for the Interior Plateau of British Columbia, where both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but where an impressive set of glacial delineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.

The Interior Plateau lineation swarm could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that the modelled LGM ice sheet is too thick, or
that the ice divide is positioned too far to the east \todo{descript the
problem for the incongruent results of these two conditions}. Because feedback
mechanisms between ice sheet topography and regional climate are absent in our
model, including wind redirection, orographic precipitation effects, and latent
warming of moisture-depleted air, we regard misfeeds in ice thickness and ice
devide location as plausible \citep{Seguinot.etal.2014}. Another explanation
for the incongruent results could be that the Interior Plateau lineation swarm
predates deglaciation and that deglacitation landforms are largely absent. The
modelled deglaciation of the Interior Plateau, central British Columbia,
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
in-between deglaciated (ice-free) mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the LGM, and may have remained intact throughout the deglaciation.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glacial maxima during
MIS~4 (61.9--55.4\,ka) and MIS~2 (29.5--16.9\,ka), two periods
corresponding to documented extensive glaciations. This result is
independent of the choice of the palaeo-temperature record used to approximate
the past climate evolution, and thus
can be seen as a first-order agreement between the model and the geological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the LGM is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by the other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the LGM.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle, confirming previous inferences from the
geological evidence preceding the LGM. However, our
simulations produce persistent ice cover on the Skeena Mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the LGM readvance, and appears
consistent with the distinct glacial imprint of the Skeena Mountains landscape.

During deglaciation, none of the palaeo-temperature records used produces a
close agreement between the model results and the geological evidence. Although
the EPICA record yields a more realistic timing of the LGM and early
deglaciation, only the GRIP record produces a late-glacial readvance in areas
where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA records, and show a rapid deglaciation of the southern
half of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. This is followed by an opening of the ice
margin in the Liard Lowland, and a final retreat of the margin of the remaining
ice caps towards the Selwyn and, later, the Skeena mountains, which host the
last remnant of the ice sheet
during the early Holocene (10.9--9.5\,kyr).

These results are strongly dependent on selected ice-sheet model (PISM),
surface
mass balance model (PDD) and climate forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, potentially leading to overestimated glacial
extent and volume in continental regions. Nevertheless, our results identify
the Skeena Mountains as a key area to understanding
glacial dynamics of the Cordilleran ice sheet, highlighting the need for
further geological investigation of this region.

@@

% Acknowledgements
\begin{acknowledgements}
We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
providing constant help and development on PISM. This work was supported by the
Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot.
Computer resources were provided by the Swedish National
Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
at the National Supercomputing Center (NSC).
\end{acknowledgements}
